Distinguishing Spin-Aligned and Isotropic Black Hole Populations With Gravitational Waves

### 2017-06-29 paper/AlignedIsotropicGW.tex
Removed footnote 1, incorporated others into text.

\documentclass{nature-arxiv}

@@

\begin{abstract}
  The first direct detections of gravitational
  waves\cite{2016PhRvL.116f1102A,2016PhRvL.116x1103A,O1-BBH,PhysRevLett.118.221101}
  from merging binary black holes open a unique window into the binary
  black hole formation environment.  One promising environmental
  signature is the angular distribution of the black hole spins;
  systems formed through dynamical interactions among already-compact
  objects are expected to have isotropic spin
  orientations\cite{SigurdssonHernquist:1993,PZMcMillan:2000,Rodriguez:2015,Stone:2016,2016ApJ...832L...2R}
  whereas binaries formed from pairs of stars born together are more
  likely to have spins preferentially aligned with the binary orbital
  angular
  momentum\cite{TutukovYungelson:1993,2016Natur.534..512B,Stevenson:2017,MandeldeMink:2016,Marchant:2016}.
  We consider existing gravitational wave measurements of the binary
  effective spin, the best-measured combination of spin
  parameters\cite{O1-BBH,PhysRevLett.118.221101}, in the four likely
  binary black hole detections GW150914, LVT151012, GW151226, and
  GW170104.  If binary black hole spin magnitudes extend to high
  values we show that the data exhibit a $\OOneSigmaIsoAligned\sigma$
  ($\OOneOddsIsoAligned$ odds ratio\footnote{An odds ratio of $r$ with
    $r \ll 1$ is equivalent to $x \sigma$ with
    $x = \Phi^{-1}\left( 1 - r/2 \right)$, where $\Phi$ is the unit
    normal CDF.})  preference for an isotropic angular distribution
  over an aligned one.  By considering the effect of 10 additional
  detections\cite{2016ApJ...833L...1A}, we show that such an augmented
  data set would enable in most cases a preference stronger than
  $5\sigma$ ($2.9 \times 10^{-7}$ odds ratio).  The existing
  preference for either an isotropic spin distribution or low spin
  magnitudes for the observed systems will be confirmed (or
  overturned) confidently in the near future.
\end{abstract}

@@

Estimates of the rate of binary black hole coalescences give a
reasonable chance of 10 additional binary black hole detections in the
next three years\cite{O1-BBH,2016ApJ...833L...1A}.  Assuming 10
additional detections drawn from each of our six zero-parameter models
in addition to the four existing detections from O1 and GW170104, with
observational uncertainties drawn randomly from the three Gaussian
widths used to approximate the $\chieff$ posteriors in Figure
\ref{fig:O1-posteriors}\footnote{The measurement uncertainty in
  $\chieff$ depends on the other parameters of the merging binary
  black hole system, particularly on the mass ratio.  Our assumption
  about future observational uncertainties is appropriate if the
  parameters of the three detected events are representative of the
  parameters of future detections. See Methods Section
  \ref{sec:chi-eff-precision} for further discussion.}, we find the
odds ratios shown in Figure \ref{fig:O2-predictions}.  We find that
most scenarios with an additional 10 detections allow the simulated
angular distribution to be inferred with greater than $5\sigma$
($2.9 \times 10^{-7}$ odds) credibility. In the most pessimistic case
the distinction is typically $\OTwoSigmaIsoAlignedMin\sigma$
($\OTwoOddsIsoAlignedMin$ odds ratio).  While such future detections
should permit a confident distinction between \emph{angular}
distributions, we would remain much less certain about the
\emph{magnitude} distribution among the three options considered here
until we have a larger number of observations.

@@

For this figure a detected population\footnote{We qualify a system as
  ``detected'' if it produces a SNR above $8$ in the second-loudest
  detector to select only coincident events.} of 500 binaries was
selected from a population with component masses distributed uniformly
between $1$ and $30~\mathrm{M}_\odot$ with
$m_1 + m_2 < 30~\mathrm{M}_\odot$, locations distributed uniformly in
volume, and orientations distributed isotropically.  Data were
simulated for each binary, and posteriors were estimated using the
LIGO-Virgo parameter estimation library
\texttt{LALInference}\cite{2015PhRvD..91d2003V} using inspiral-only
waveform models (merger and ringdown effects can provide additional
information for some binaries, but we ignore those effects here).
$\chieff$ is better constrained for binaries with high
effective spins and high ($\sim$equal) mass ratios.

>>>>

\documentclass{nature}

@@

\begin{abstract}
  The first direct detections of gravitational
  waves\cite{2016PhRvL.116f1102A,2016PhRvL.116x1103A,O1-BBH,PhysRevLett.118.221101}
  from merging binary black holes open a unique window into the binary
  black hole formation environment.  One promising environmental
  signature is the angular distribution of the black hole spins;
  systems formed through dynamical interactions among already-compact
  objects are expected to have isotropic spin
  orientations\cite{SigurdssonHernquist:1993,PZMcMillan:2000,Rodriguez:2015,Stone:2016,2016ApJ...832L...2R}
  whereas binaries formed from pairs of stars born together are more
  likely to have spins preferentially aligned with the binary orbital
  angular
  momentum\cite{TutukovYungelson:1993,2016Natur.534..512B,Stevenson:2017,MandeldeMink:2016,Marchant:2016}.
  We consider existing gravitational wave measurements of the binary
  effective spin, the best-measured combination of spin
  parameters\cite{O1-BBH,PhysRevLett.118.221101}, in the four likely
  binary black hole detections GW150914, LVT151012, GW151226, and
  GW170104.  If binary black hole spin magnitudes extend to high
  values we show that the data exhibit a $\OOneSigmaIsoAligned\sigma$
  ($\OOneOddsIsoAligned$ odds ratio) preference for an isotropic
  angular distribution over an aligned one.  By considering the effect
  of 10 additional detections\cite{2016ApJ...833L...1A}, we show that
  such an augmented data set would enable in most cases a preference
  stronger than $5\sigma$ ($2.9 \times 10^{-7}$ odds ratio).  The
  existing preference for either an isotropic spin distribution or low
  spin magnitudes for the observed systems will be confirmed (or
  overturned) confidently in the near future.
\end{abstract}

@@

Estimates of the rate of binary black hole coalescences give a
reasonable chance of 10 additional binary black hole detections in the
next three years\cite{O1-BBH,2016ApJ...833L...1A}.  Assuming 10
additional detections drawn from each of our six zero-parameter models
in addition to the four existing detections from O1 and GW170104, with
observational uncertainties drawn randomly from the three Gaussian
widths used to approximate the $\chieff$ posteriors in Figure
\ref{fig:O1-posteriors}, we find the odds ratios shown in Figure
\ref{fig:O2-predictions}.  The measurement uncertainty in $\chieff$
depends on the other parameters of the merging binary black hole
system, particularly on the mass ratio.  Our assumption about future
observational uncertainties is appropriate if the parameters of the
three detected events are representative of the parameters of future
detections. See Methods Section \ref{sec:chi-eff-precision} for
further discussion.  We find that most scenarios with an additional 10
detections allow the simulated angular distribution to be inferred
with greater than $5\sigma$ ($2.9 \times 10^{-7}$ odds)
credibility. In the most pessimistic case the distinction is typically
$\OTwoSigmaIsoAlignedMin\sigma$ ($\OTwoOddsIsoAlignedMin$ odds ratio).
While such future detections should permit a confident distinction
between \emph{angular} distributions, we would remain much less
certain about the \emph{magnitude} distribution among the three
options considered here until we have a larger number of observations.

@@

For this figure a detected population of 500 binaries was selected
from a population with component masses distributed uniformly between
$1$ and $30~\mathrm{M}_\odot$ with $m_1 + m_2 < 30~\mathrm{M}_\odot$,
locations distributed uniformly in volume, and orientations
distributed isotropically.  We qualify a system as ``detected'' if it
produces a SNR above $8$ in the second-loudest detector to select only
coincident events.  Data were simulated for each binary, and
posteriors were estimated using the LIGO-Virgo parameter estimation
library \texttt{LALInference}\cite{2015PhRvD..91d2003V} using
inspiral-only waveform models (merger and ringdown effects can provide
additional information for some binaries, but we ignore those effects
here).  $\chieff$ is better constrained for binaries with high
effective spins and high ($\sim$equal) mass ratios.
